---
title: "HW-01"
author: "Sterling Hayden"
date: "2024-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(earth)
library(ROCit) #ROC curve
library(mgcv)
```

```{r}
df_train = read.csv('insurance_t.csv')
df_val = read.csv('insurance_v.csv')
```

### Checking for missingness
```{r}
# Find columns with missing data
miss_t <- names(df_train)[colSums(is.na(df_train)) > 0]
miss_v <- names(df_val)[colSums(is.na(df_val)) > 0]

# Check if train and val are missing the same data
print(identical(miss_t, miss_v))

# Check what and how many values are missing in the train dataset
colSums(is.na(df_train[miss_t]))

```

Impute median for continous varriables and a missing flag for the categorical columns
```{r}
# Define columns
continuous_cols <- c("ACCTAGE", "PHONE", "POS", "POSAMT", "INVBAL", "CCBAL", "INCOME", "LORES", "HMVAL", "AGE", "CRSCORE")
binary_cols <- c("INV", "CC", "CCPURC")

# Replace 'ns' values with 'Missing', 1 with 'Instance', and 0 with 'Non-Instance' for train and val datasets
for (col in binary_cols) {
  df_train[[col]] <- ifelse(is.na(df_train[[col]]), "Missing",
                            ifelse(df_train[[col]] == 1, "Instance", "Non-Instance"))
  df_val[[col]] <- ifelse(is.na(df_val[[col]]), "Missing",
                          ifelse(df_val[[col]] == 1, "Instance", "Non-Instance"))
}

# Impute with median and create binary flag column
for (col in continuous_cols) {
  median_value <- median(df_train[[col]], na.rm = TRUE)
  # Create a binary flag variable for imputation in train dataset
  df_train[[paste0(col, "_imputed")]] <- ifelse(is.na(df_train[[col]]), 1, 0)
  # Impute the missing values with the median in train dataset
  df_train[[col]][is.na(df_train[[col]])] <- median_value
}

# Repeat the same process for the validation dataset
for (col in continuous_cols) {
  median_value <- median(df_train[[col]], na.rm = TRUE)
  # Create a binary flag variable for imputation in val dataset
  df_val[[paste0(col, "_imputed")]] <- ifelse(is.na(df_val[[col]]), 1, 0)
  # Impute the missing values with the median in val dataset
  df_val[[col]][is.na(df_val[[col]])] <- median_value
}
```



### MARS/EARTH algorithm

```{r}
mars1 <- earth(INS ~ ., data = df_train, glm = list(family = binomial))
summary(mars1)
```

Varriable Importance
```{r}
evimp(mars1)
```
The more subsets (or “best models”) that a variable appears in, the more important the variable. The most import variable is SAVBAL, then CDBAL, and so on. 


ROC Curve
```{r}
df_train$p_hat_EARTH <- predict(mars1, type = "response")
roc_earth <- rocit(score = as.vector(df_train$p_hat_EARTH), class = df_train$INS)
plot(roc_earth)
plot(roc_earth)$optimal
paste("AUC =",roc_earth$AUC)
```




### GAM algorithm

```{r}
cols_with_more_than_10 <- names(df_train)[sapply(df_train, function(col) n_distinct(col) > 10)]
cols_with_less_than_10 <- names(df_train)[sapply(df_train, function(col) n_distinct(col) <= 10)]
```


```{r}
gam1 <- mgcv::gam(INS ~ 
                  s(ACCTAGE) + 
                  s(DDABAL) +
                  s(DEP) +
                  s(DEPAMT) +
                  s(CHECKS) +
                  s(NSFAMT) +
                  s(PHONE) +
                  s(TELLER) +
                  s(SAVBAL) +
                  s(ATMAMT) +
                  s(POS) +
                  s(POSAMT) +
                  s(CDBAL) +
                  s(IRABAL) +
                  s(INVBAL) +
                  s(MMBAL) +
                  s(CCBAL) +
                  s(INCOME) +
                  s(LORES) +
                  s(HMVAL) +
                  s(AGE) +
                  s(CRSCORE) +
                  factor(DDA) +
                  factor(DIRDEP) +
                  factor(NSF) +
                  factor(SAV) +
                  factor(ATM) +
                  factor(CD) +
                  factor(IRA) +
                  factor(INV) +
                  factor(MM) +
                  factor(MMCRED) +
                  factor(CC) +
                  factor(CCPURC) +
                  factor(SDB) +
                  factor(INAREA) +
                  factor(ACCTAGE_imputed) +
                  factor(PHONE_imputed) +
                  factor(POS_imputed) +
                  factor(POSAMT_imputed) +
                  factor(INVBAL_imputed) +
                  factor(CCBAL_imputed) +
                  factor(INCOME_imputed) +
                  factor(LORES_imputed) +
                  factor(HMVAL_imputed) +
                  factor(AGE_imputed) +
                  factor(CRSCORE_imputed) +
                  factor(BRANCH),
                  family = binomial, method = 'REML', data = df_train)

save(gam1, file = "gam1.RData")
load("gam1.RData")
summary(gam1)
```

penalized GAM w/ select = TRUE
```{r}
sel.gam2 <- mgcv::gam(INS ~ 
                  s(ACCTAGE) + 
                  s(DDABAL) +
                  s(DEP) +
                  s(DEPAMT) +
                  s(CHECKS) +
                  s(NSFAMT) +
                  s(PHONE) +
                  s(TELLER) +
                  s(SAVBAL) +
                  s(ATMAMT) +
                  s(POS) +
                  s(POSAMT) +
                  s(CDBAL) +
                  s(IRABAL) +
                  s(INVBAL) +
                  s(MMBAL) +
                  s(CCBAL) +
                  s(INCOME) +
                  s(LORES) +
                  s(HMVAL) +
                  s(AGE) +
                  s(CRSCORE) +
                  factor(DDA) +
                  factor(DIRDEP) +
                  factor(NSF) +
                  factor(SAV) +
                  factor(ATM) +
                  factor(CD) +
                  factor(IRA) +
                  factor(INV) +
                  factor(MM) +
                  factor(MMCRED) +
                  factor(CC) +
                  factor(CCPURC) +
                  factor(SDB) +
                  factor(INAREA) +
                  factor(ACCTAGE_imputed) +
                  factor(PHONE_imputed) +
                  factor(POS_imputed) +
                  factor(POSAMT_imputed) +
                  factor(INVBAL_imputed) +
                  factor(CCBAL_imputed) +
                  factor(INCOME_imputed) +
                  factor(LORES_imputed) +
                  factor(HMVAL_imputed) +
                  factor(AGE_imputed) +
                  factor(CRSCORE_imputed) +
                  factor(BRANCH),
                  family = binomial, method = 'REML', select = TRUE, data = df_train)

save(sel.gam2, file = "sel_gam2.RData")
summary(sel.gam2)
```


Building a GAM w/ only the p-values > 0.05 from the full model
```{r}
final.gam <- mgcv::gam(INS ~ 
                        s(ACCTAGE) + 
                        s(DDABAL) + 
                        s(CHECKS) +
                        s(TELLER) + 
                        s(SAVBAL) +
                        s(ATMAMT) +
                        s(CDBAL) +
                        s(CCBAL) + 
                        factor(DDA) +
                        factor(NSF) +
                        factor(CD) + 
                        factor(IRA) + 
                        factor(INV) +
                        factor(MM) +
                        factor(CC) + 
                        factor(BRANCH),
                  family = binomial, method = 'REML', data = df_train)

save(final.gam, file = "final_gam.RData")
summary(final_gam)
```


ROC Curve
```{r}
df_train$p_hat_GAM <- predict(gam1, type = "response")
roc_gam <- rocit(score = as.vector(df_train$p_hat_GAM), class = df_train$INS)
plot(roc_gam)
plot(roc_gam)$optimal
paste("AUC =",logit_roc$AUC)
```
final.gam has AUC = 0.801639776016325
sel.gam2 has  AUC = 0.800576402912038
gam1 has      AUC = 0.80326497635508
earth/mars    AUC = 0.799357379760185


```{r}
# Plot the first ROC curve
plot(roc_earth, col = "blue", main = "Overlayed ROC Curves", lwd = 2, legend = FALSE)

# Overlay the second ROC curve using the lines() function
lines(roc_gam$TPR ~ roc_gam$FPR, col = "red", lwd = 2)

# Add a legend to distinguish the curves
legend("bottomright", legend = c("EARTH Model", "Final/Hand Selected GAM Model"), col = c("blue", "red"), lwd = 2)


```

```{r}
# Extract the p-values for the smooth terms
summary_gam <- summary(gam1)
p_values_smooth <- summary_gam$s.table[, 4]

# Sort the p-values in ascending order for the smooth terms
sorted_p_values_smooth <- sort(p_values_smooth)

# Print the sorted smooth term p-values
print(sorted_p_values_smooth)

# If you have parametric terms and want to sort them as well:
p_values_parametric <- summary_gam$p.table[, 4]  # Assuming the 4th column has the p-values

# Sort the p-values for the parametric terms
sorted_p_values_parametric <- sort(p_values_parametric)

# Print the sorted parametric term p-values
print(sorted_p_values_parametric)

```
